---
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# NestWatch Data Products

<!-- badges: start -->

![Static Badge](https://img.shields.io/badge/Dataset%20License-CC%20BY--NC%204.0-blue)\
![Static Badge](https://img.shields.io/badge/Code%20License-GPL%20v3-blue)

<!-- badges: end   -->

## Development Tasks :construction:

This repository is in early development with plans to add additional functions to create an R package to facilitate accessibility.

   🔲 Modify nw.filterphenology() to more properly handle run-on nesting attempts

   🔲 Finish phenological filtering example in "filter" vignette

   🔲 Add reproducible example short code to function documentations where missing
   
   🔲 Internal testing

    

## Overview :pushpin:

[NestWatch](http://www.nestwatch.org) is a community science project operated by the Cornell Lab of Ornithology. Since 1965, members of the public have been following a standardized protocol for observing and reporting birds' nests in the United States and Canada (and more recently, globally). This dataset contains raw nest records submitted to NestWatch. The dataset contains millions of nest check observations from hundreds of thousands of nest attempts. The purpose of this repository is to provide a collection of functions to aid in accessing and analyzing the wealth of nesting data contained in the NestWatch database.

**The metadata paper associated with this dataset [(Bailey et al. 2023)](https://doi.org/10.1002/ecy.4230) is critical for understanding and interpreting data fields and their contents**. This dataset is scheduled for updates annually on or about January 31. Relevant details can also be found on the [NestWatch website here](https://nestwatch.org/explore/nestwatch-open-dataset-downloads/).

## Installation :computer:

This package is under active development and is likely to have bugs and unexpected issues. Presently, the package exists off-CRAN in the public `engagement-center/nestwatchR` GitHub repository. To install the package on your local machine, either run the following code, or clone the repository to your local machine and use the "build" tab of the environments pane to install the package, for testing.

```{r install remotes, eval=FALSE, meaasge=F}
# Install nestwatchR package in R
remotes::install_github("engagement-center/nestwatchR")
```

This package primarily uses the `R` programming language, and some knowledge of R is necessary for using these products. A function in this collection also depends on the program `Python`, which needs to be [installed](https://www.python.org/downloads/) on your machine prior to using the function. No prior knowledge of Python or manual setup of the program is needed. Use the following code chunk to check if Python is already installed on your machine and follow the resulting link if not already installed:

```{r gh-installation, eval = FALSE}
# Check to see if you already have Python installed
check_python_installation <- function() {
  python_installed <- system("python --version", intern = TRUE)
  if (length(python_installed) > 0) {
    print("Python is installed.")
  } else {
    message("Python is not installed.")
    message("Please download and install Python from https://www.python.org/downloads/")
  }
}
check_python_installation()
```

These functions also depend on several `R` libraries. If you do not already have these packages installed, we suggest installing them during the `nestwatchR` installation process.

```{r package dependencies, evel = FALSE}
# List of packages used in NestWatch code
packages <- c("reticulate", "dplyr", "stringr", "lubridate", "ebirdst", "sf")

# Check if each package is installed, and if not, install it
for (pkg in packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# Load required libraries
library(dplyr); library(stringr); library(lubridate); library(ebirdst); library(sf); library(reticulate)

```

## Data access :globe_with_meridians:

Data from NestWatch is free to access either from the `nw.getdata` function in R or as a direct download from the [Mendeley Data Archive](https://data.mendeley.com/datasets/wjf794z7gc/2) or NestWatch website. NestWatch Open Dataset by Cornell Lab of Ornithology is licensed under CC BY-NC 4.0. We encourage researchers seeking to conduct formal analyses to use these data. As with any dataset, knowing the data structure, understanding the metadata, grasping the data collection protocols, and being aware of the unique aspects of the program are all critical for conducting analyses and interpreting results in ways that provide meaningful insights. Prior to analysis, we encourage all users of NestWatch Data to read the article ["Download Raw NestWatch Data for Analysis"](https://nestwatch.org/explore/nestwatch-open-dataset-downloads/) article on our website.

## Versions :memo:

This collection of functions is on Version 0.0 and will be updated to maintain functionality and improve accessibility to the NestWatch datasets.

The NestWatch Open Dataset is updated annually, by or around 31 January, with the previous year's nesting records. NestWatch also updates the database with historical nest records being archived from other projects. To improve reproducibility in NestWatch analyses, each version of the dataset is archived in Mendeley, and each specific version can be downloaded either using the `nw.getdata(version = )` function or by a direct download from the [Mendeley archive](https://data.mendeley.com/datasets/wjf794z7gc/2).

## Citation :book:

If you use the NestWatch Open Dataset or this R package, please cite:

<blockquote>

**Dataset (replace with appropriate version \# and year):**</br> Bailey, R., L. Larson, D. Bonter. 2024. “NestWatch Open Dataset." Mendeley Data, V2. doi: 10.17632/wjf794z7gc.2</a>

**Data Paper:**</br> Bailey, R. L., L. Larson, and D. N. Bonter. 2024. “NestWatch: An Open-Access, Long-Term Data Set on Avian Reproductive Success.” Ecology 105(2): e4230. https://doi.org/10.1002/ecy.4230</a>

**NestWatch R Package:**<br> Savides, K., R. Bailey, & D. Bonter. NestWatch Data Products (Version 0.00.00) [Computer software]. <https://github.com/engagement-center/2024-NestWatch-Package-Internal>

</blockquote>

## Vignettes :books:

For full function documentation, including a series of vignettes covering introductory usage of NestWatch data, please see the RMarkdown Vignettes:

-   [Introduction to NestWatch Data and Data Access](https://engagement-center.github.io/nestwatchR/articles/a_Into-and-Data-Access.html): covers data access, available data products, and introduction to structure and format of data files
-   [Conduct Common NestWatch Data Cleaning Procedures](https://engagement-center.github.io/nestwatchR/articles/b_Data-Cleaning.html): demonstrates how and when to use a variety of common data cleaning procedures designed for NestWatch data
-   [Filter NestWatch Data on Finer Scales](https://engagement-center.github.io/nestwatchR/articles/c_Data-Filtering.html): demonstrates the use of species-level data filters and functions to estimate missing values
-   [Estimate NestWatch Summary Dates](https://engagement-center.github.io/nestwatchR/articles/d_Data-Estimation.html): demonstrates functions to estimate missing summary values

## Quick Start :ledger:

This quick start guide shows how to download data and plot the first lay dates of two species. In this guide, and throughout all package documentation, a simplified example dataset is used consisting of Carolina and Bewick's wrens.

```{r background operations, include = FALSE, echo = FALSE}
library(nestwatchR)
load("data/wren_quickstart.rda")
wrens <- wren_quickstart
rm(wren_quickstart)
```

```{r code but dont run, eval=FALSE}
# Load NestWatch Package
library(nestwatchR)

# Download NestWatch dataset by version
nw.getdata(version = 2)

# Merge the Attempts and Checks files
nw.mergedata(attempts = NW.attempts, checks = NW.checks, output = "merged.data")

nrow(merged.data)
#>  [1] 2639824

# Filter the dataset to include just Carolina and Bewick's wrens
wrens <- merged.data %>% filter(Species.Code %in% c("carwre", "bewwre"))
nrow(wrens)
#>  [1] 40290
head(wrens, 5)
```

```{r head of wren DO NOT MODIFY its manually formatted, max.height='100px', max.width='100px'}
#> X  Attempt.ID  Location.ID   Latitude    Longitude Subnational.Code  Species.Name  Species.Code  Year  Elevation.m  Height.m Substrate Substrate.Relationship  Substrate.Other.Description Predator.Guard  Predator.Guard.Other  Cavity.Entrance.Diameter.cm Entrance.Orientation  Habitat.1m  Habitat.100m  Location.Entry.Technique  Observer.ID First.Lay.Date  First.Lay.Date.Estimated  Visited.During.Egg.Laying Hatch.Date  Hatch.Date.Estimated  Fledge.Date Fledge.Date.Estimated Young.Fledged Clutch.Size Young.Total Unhatched.Eggs  Outcome Attempt.Entry.Technique   Visit.ID       Visit.Datetime  Visit.Time.Valid  Nest.Status Adult.Activity  Young.Status  Management.Status Host.Eggs.Count Host.Eggs.Present.Uncounted Live.Host.Young.Count Live.Host.Young.Present.Uncounted Dead.Host.Young.Count Dead.Host.Young.Present.Uncounted Brood.Parasite.Eggs.Count Brood.Parasite.Eggs.Present.Uncounted Brood.Parasite.Live.Young.Count Brood.Parasite.Live.Young.Present.Uncounted Brood.Parasite.Dead.Young.Count Brood.Parasite.Dead.Young.Present.Uncounted Visit.Entry.Technique 
#> 1	A1000045	   L56654	34.68472	-98.40583	         US-OK	Carolina Wren	    carwre	2006	    344.7	 1.6002	   nesbox	                  in		                                               		                               3.81	                  ne 	               	             BIRDHOUSE_HIST:SELF     obsr1309		                                NA	                       NA		                        NA				                   NA	         NA	         NA          NA	            NA	      f	                       	  S2489482	2006-07-10 00:00:00	                0	                   			                               	             5                           0                    NA                                 0                    NA                                 0                        NA                                     0                              NA                                           0                              NA                                           0
#> 2	A1000045	   L56654	34.68472	-98.40583	         US-OK	Carolina Wren	    carwre	2006	    344.7	 1.6002	   nesbox	                  in		                                               		                               3.81	                  ne 	               	             BIRDHOUSE_HIST:SELF     obsr1309		                                NA	                       NA		                        NA				                   NA	         NA	         NA          NA	            NA	      f	                       	  S2489483	2006-08-08 00:00:00	                0					       no	                            nm              NA                           0                    NA                                 0                    NA                                 0                        NA                                     0                              NA                                           0                              NA                                           0   
#> 3	A1000271	  L126818	30.61476	-76.87363	         US-TX	Bewick's Wren	    bewwre	2006	    316.4	 1.6154	   nesbox	                  in		                              baffle	                                           3.81	                  sw 	               	                 PointMaker1.0_2    obsr10027		                                NA	                       NA		                        NA				                   NA	         NA	         NA          NA	            NA	     f6	                          S2494112	2006-03-16 00:00:00	                0	        in                                                              NA                           0                    NA                                 0                    NA                                 0                        NA                                     0                              NA                                           0                              NA                                           0   
#> 4	A1000271	  L126818	30.61476	-76.87363	         US-TX	Bewick's Wren	    bewwre	2006	    316.4	 1.6154	   nesbox	                  in		                              baffle	                                           3.81	                  sw 	               	                 PointMaker1.0_2    obsr10027		                                NA	                       NA		                        NA				                   NA	         NA	         NA          NA	            NA	     f6	                          S2494113	2006-03-13 00:00:00	                0	        cn                                                              NA                           0                    NA                                 0                    NA                                 0                        NA                                     0                              NA                                           0                              NA                                           0   
#> 5	A1000271	  L126818	30.61476	-76.87363	         US-TX	Bewick's Wren	    bewwre	2006	    316.4	 1.6154	   nesbox	                  in		                              baffle	                                           3.81	                  sw 	               	                 PointMaker1.0_2    obsr10027		                                NA	                       NA		                        NA				                   NA	         NA	         NA          NA	            NA	     f6	                          S2494114	2006-04-12 00:00:00	                0	                                                                         4                           0                    NA                                 0                    NA                                 0                        NA                                     0                              NA                                           0                              NA                                           0   
```

```{r plot, eval =TRUE}
# Format First.Lay.Date as a date
wrens$First.Lay.Date <- as.Date(wrens$First.Lay.Date)

# Filter out NA values of First Lay Date, force all attempt years to 2024 for visualization
data <- wrens %>% filter(!is.na(First.Lay.Date)) %>% 
                  mutate(First.Lay.Date = update(First.Lay.Date, year = 2024))
  
# Plot the data in ggplot
library(ggplot2)
ggplot(data, aes(x = First.Lay.Date, fill = Species.Name)) +
  geom_histogram(binwidth = 7, alpha = 0.5, position = "stack", color = "black") + 
  labs(x = "First Lay Date", y = "Count", title = "First Lay Date by Species") +
  theme_minimal()

```
