---
title: "Conduct Common NestWatch Data Cleaning Procedures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Coduct Common NestWatch Data Cleaning Procedures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      collapse = TRUE,
                      comment = "#>",
                      out.width = "100%",
                      fig.height = 4, 
                      fig.width = 7, 
                      fig.align = "center")
# only build vignettes locally and not for R CMD check
knitr::opts_chunk$set(eval = nzchar(Sys.getenv("BUILD_VIGNETTES")))
library(nestwatchR)
```

This vignette will cover several different methods for cleaning data found within the NestWatch database. NestWatch data are collected by volunteer participants (researchers & the public) and are known to contain some common errors and inconsistencies. Thus, these data should be considered as "raw" and the user should consider their individual study system and research objectives when deciding what cleaning procedures to conduct.This package contains two function to provide coarse-scale cleaning procedures that NestWatch staff have identified in order to help the user clean NestWatch data for analysis. The user should read the full function documentation and this vignette prior to cleaning the data.

# Background

NestWatch data entered in real-time from the U.S. and Canada are passed through a series of filters that attempt to catch data-entry errors or prevent unlikely responses in clutch size, number of young, timing of nesting, incubation periods, and brooding periods. This real-time feedback system does not identify all potential errors. For instance, if a participant enters the number of eggs incorrectly but the typo in clutch size is still biologically possible, no feedback will be given. Even if real-time feedback is given, the participant may override the warning if they feel the entry they have made is accurate. Participants do occasionally observe unusually large clutch sizes, or nests that have occurred well outside of the traditional nesting period for the focal species. It is up to each analyst to decide if these observations are relevant. Importantly, these initial biological filters have not been established for species nesting outside of the U.S. and Canada or for species that are seldom reported.

NestWatch additionally implements geographic filters on real-time entered species, which are curated by eBird. In other words, a participant should not be able to report a nest from a species that is extremely rare in that region; thus, many species misidentifications are prevented. However, for similar sympatric species, incorrect identifications are still possible (e.g., a Black-capped Chickadee may be misidentified as a Carolina Chickadee in regions of range overlap). However, older reports, including historic data migrated to the NestWatch database, not subjected to the same rigorous filters. Additionally, errors in reporting can mimic errors in identification. Participants may intend to report one species but enter their information for the wrong species. We recommend that data analysts map nest locations and screen for likely errors. See the `nw.filterspatial()` and Filtering vignettes for more information on spatial filtering. 

# Clean Taxa

The NestWatch platform uses [eBird taxonomy](https://science.ebird.org/en/use-ebird-data/the-ebird-taxonomy) to help users auto-complete species names. This allows users to select any available eBird taxonomic unit, including "spuhs" (ie. "chickadee sp."), slashes (ie. "Tree/Violet-green Swallow"), hybrids (ie. "Mallard x American Black Duck (hybrid)"), and subspecies/forms (ie."House Wren (northern)"). `nw.cleantaxa` helps a user filter out and/or "roll up" these taxonomic units. This function uses a NestWatch dataframe as `data =` and has logical arguments `spuh =`, `slash =`, `hybrid =`, and `rollsubspecies =` where `T` removes attempts with those designations and subspecies/forms are "rolled up" to the species level (ie. "House Wren (northern)" -> "House Wren"). If arguments are set to `F` or `NULL` the filtering and/or roll up are not preformed. The resulting dataframe may be optionally names with `output =`. 

This function changes both the common name and the species code to reflect the changes made to the taxonomic unit. Taxonomy updates are implemented via `auk::ebird_taxonomy` and are updated with the annual eBird taxonomic update. For more information, see [`auk::ebird_taxonomy`](https://cornelllabofornithology.github.io/auk/reference/ebird_taxonomy.html) and [eBird Taxonomy](https://science.ebird.org/en/use-ebird-data/the-ebird-taxonomy). 

```{r cleantaxa 1, eval = T, echo = T}
# Simple example dataframe
df <- data.frame(Attempt.ID = c(1, 1, 2, 3, 4),
                 Species.Name = c("House Wren (northern)", "House Wren (northern)",
                                 "chickadee sp.", "Tree/Violet-green Swallow",
                                 "Mallard x American Black Duck (hybrid)"),
                 Species.Code = c("houwre1", "houwre1", "chicka1", "y00701", "x00004"))
df
```

### Sphus and Slashes

Correct usage of a spuh would, for example, be if a participant observed a chickadee nesting in a region of Black-capped/Carolina Chickadee range overlap and was unable to confirm the exact species with certainty. The same would be true if a participant monitored a nest box of a swallow species and never saw the adults at a site where both Tree and Violet-green Swallows were present. In both cases, either a slash or spuh could be appropriate. 

When certainty of a specie(s) is needed, the user may decide to omit spuhs and slashes as species is not known with certainty:


```{r cleantaxa 2, eval = T, echo = T}
# Remove spuhs and slashes
nw.cleantaxa(data = df, spuh = T, slash = T, output = "ex1")
ex1
```

### Subspecies/forms & Hybrids

Some subspecies or forms are easy differentiated and may be specified by a participant. The user should use caution when using these subspecies/form designations as the identification skills of the participant are unknown and the participant may not have selected the correct designation for the populated list of possible species. The user should also consider if there may or may not be biologically relevant differences in nesting behavior, phenology, or other parameter between different subspecies/forms before cleaning or analyzing these data. If these potential differences are not biologically relevant to the analysis, the user may use `rollsubspecies = T` to combine forms to the species level.

Documentation of nesting attempts between different parent species, a hybrid parent, or two hybrid parents are very uncommon. The NestWatch database has no way to easily differentiate between these three cases. Thus if a hybrid species was selected for an attempt, the parent specie(s) of the nest vs the resulting young specie will not be known with certainty. Additionally, NestWatch staff have confirmed in some cases that participants incorrectly selected a hybrid species when they should have selected a spuh or slash species designation. If a user is interested in validating attempts labeled with hybrid status, they may [request additional information](https://data.nestwatch.org/nw/contact) from NestWatch staff from "notes" fields which *may contain additional information not exported in the public database*. 


```{r cleantaxa3, eval = T, echo = T}
# Remove hybrids and roll up subspecies
nw.cleantaxa(data = df, hybrid = T, rollsubspecies = T, output = "ex2")
ex2
```



# Clean Data

The function `nw.cleandata()` is a collection of 10 common cleaning procedures which NestWatch staff have identified to assist users in cleaning NestWatch data at a coarse-scale. Users need to specify which procedures, lettered `a:j`, to run and should read the function documentation and the below examples while considering their analyses before selecting which procedures to run. After these coarse-scale cleanings are completed, the user should continue cleaning the data at a finer-scale including considering spatial and phonological filtering. See the filtering vignette and `nw.filterspatial()` and `nw.filterphenology()` functions.

`nw.cleandata()` takes a NestWatch dataframe as `data =` and allows the user to specify the function's `mode = ` as either `"flag"` or `"remove"`. When mode is set to flag, nest attempts which do not meet the cleaning criteria will be noted in a new column `Flagged.Attempt` with the value `FLAGGED` or `NA`. If mode is set to remove, these flagged attempts are removed from the output. 

> ![NOTE]
> If you are only interested in a subset of species, or RAM usage and processing time are a concern, you may choose to subset the data to those species prior to running `nw.cleandata`.  

```{r cleandata 1, eval = F, echo = T}
# If interested only in a few species, subset to example of Carolina and Bewick's Wrens
# Using output from `nw.mergedata`
df <- merged.data %>% filter(Species.Code %in% c("carwre", "bewwre"))

# Clean data function parameters
nw.cleandata(data = df, mode = "flag", methods = c("a", "b", "c") , output = "cleaned.data")

```

Individual cleaning procedures are specified in a character vector of lowercase letters `a:j`. An optional `ouptput =` argument is included to rename the resulting dataframe as well.

## Cleaning Procudure Definitions

a.  **Brown-headed Cowbirds:**  Flag/remove attempts with `Species.Name` as "Brown-headed Cowbird". Brown-headed Cowbirds are nest parasites and do not create their own nests. This is not the correct interpretation of the nest species. Users may choose to look at these data more closely if investigating brood parasitism. <br>
<br>
a.  **No Netsing Activity:**  Flag/remove attempts with Outcome of "no breeding behavior observed" (u3), "inactive" (i), and "not monitored" (n). These data likely represent either nests from a previous season or nests that were previously active this season, but no active dates can be attributed to. Users may choose to include these data if looking at habitat or location data without measures of phenology or success. <br>
<br>
a.  **Invasive spp Managment:**  Flag/remove attempts with Outcome of "invasive spp management" (f5). In these attempts, the participant chose to remove/alter eggs or nests of invasive species. If the user is interested in analyzing participants' habits in invasive spp management or rates of brood parasitism, they should consider not using this method. Note: "Invasive spp management", may also be incorrectly chosen by participants trying to indicate nest failure by an invasive species where a failure code should have been used. <br>
<br>
a.  **Failed but fledged?:**  Flag/remove attempts if outcome is a fail code (f, f1, f2, f3, f6, f7) but recorded fledged host young > 0. In this case, a participant may have either characterized the nest's outcome incorrectly, recorded the presence of fledged host young incorrectly, or mischaracterized brood parasite young as host, but discerning the true outcome may not be possible. If an attempt produces any number of fledged host young, the attempt was successful.<br>
<br>
a.  **Successful but didn't fledge?:**  Flag/remove attempt if Outcome is success (s1), but recorded fledged host young = 0. Inverse of d.<br>
<br>
a.  **More young than eggs?:**  Flag/remove attempt if # hatched host young > clutch size. This may indicate incorrectly entered data or long lengths of time between nest checks where summary data was not properly updated. A user may choose to review these attempts by looking with caution at the nest visit data to validate hatched young and clutch sizes.<br>
<br>
a.  **More fledged young than young in nest:**  Flag/remove attempt if # fledged host young > # hatched host young. Similar to f.<br>
<br>
a.  **International Waters:**  Flag/remove attempts that occur in international waters Subnational.Code == "XX-". Consider not using this procedure if analyzing coastal or water-nesting species (eg. Osprey nesting on channel markers). Subnational code is automatically assigned based on coordinates supplied by the participant. Many attempts identified as XX- likely resulted from incorrect or inaccurate coordinates.<br>
<br>
a.  **Impossibly long attempts:**  Flag attempts with impossible nesting periods. Incorrect years are sometimes entered between date summary fields by participants. This may produce impossibly long nesting periods. To account for nesting phenologies in all hemispheres, this procedure identifies attempts in which (1) Fledge Date - First Lay Date > 365 days (Snowy Albatross nesting period), (2) Hatch Date - First Lay Date > 84 days (Southern Brown Kiwi incubation period), or (3) Fledge Date - Hatch Date > 300 days (Snowy Albatross nestling period). These dates represent the maximum nest phonological period for any bird species and are not realistic for the majority of the NestWatch dataset. We encourage users to determine reasonable phenologies for their species of interest and use `nw.filterphenology()` to run a finer filter on nest phenology dates.<br>
<br>
a.  **Year errors in visits:**  Flag/remove attempts where the # days between the first and the last visit are > 365 days. Additional check to identify nest attempts where year portion of dates between nest visits were likely incorrect entered. A user may choose to review these attempts individually to verify if a typo occurred.<br>
<br>


Depending on what cleaning procedures you choose to run, the sample size of nests may decrease to varying degrees. If your species has a small sample size after using `mode = "remove"` you may consider rerunning with `mode = "flag"` and inspecting which attempts were removed and if each cleaning method was appropriate for your analysis. 

```{r cleandata ex, eval = F, echo = T}
# Load and merge version 2 of database
nw.getdata(version = 2)
nw.mergedata(attempts = NW.attempts, checks = NW.checks, output = "merged.data")
length(unique(mergeddata$Attempt.ID))
# > 572488    # Attempts in raw dataset


# Remove attempts with 'Brown-headed Cowbird' as the species
nw.cleandata(data = mergeddata, mode = "remove", methods = "a", output = "no_bnhcow")
length(unique(no_bnhcow$Attempt.ID))
# > 574263    # Attempts after cleaning

# Remove attempts using "i" and "j"
nw.cleandata(data = mergeddata, mode = "remove", methods = c("i", "j"), output = "no_longnests")
length(unique(no_longnests$Attempt.ID))
# > 572583    # Attempts after cleaning
```
