---
title: "Estimate NestWatch Summary Dates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimate NestWatch Summary Dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      collapse = TRUE,
                      comment = "#>",
                      out.width = "100%",
                      fig.height = 4, 
                      fig.width = 7, 
                      fig.align = "center")
# only build vignettes locally and not for R CMD check
knitr::opts_chunk$set(eval = nzchar(Sys.getenv("BUILD_VIGNETTES")))
library(nestwatchR)
library(dplyr)
library(DT)
```


The NestWatch database holds a wealth of nesting data. But, due to participant omissions, differing protocols over time, and assimilation of other datasets, not all of these data are in the appropriate summary data fields. For instance, a participant may forget to input the clutch size, but that data may exist in the nest visits section if the participant recorded numbers of eggs and chicks during their nest visits. The `nw.est...` functions help users estimate `Clutch.Size`, `First.Lay.Date`, `Hatch.Date`, and `Fledge.Date`. 

Each of these four estimation functions are coded to estimate based on the most proximal event to that being estimated or based off the event with the least general variability in birds (see each section below). Though these functions try to use the best available data, the precision of the resulting estimates can not be guaranteed. 

## Estimate Clutch Size
Clutch size is a common omission in the NestWatch dataset. `nw.estclutchsize()` takes a merged NestWatch dataframe as `data =`, and has an optional `output =` argument to rename the resulting dataframe. 

This function looks for data contained in several columns to estimate what the clutch size of the nest was. Because NestWatch participants do not necessarily visit each nest everyday or record all data fields on every visit the data needed to estimate clutch size may be spread over many columns in the nest visits section. The function looks, by Attempt.ID, in the nest visits and summary data and finds (1) the maximum egg count, (2) the unhatched eggs count, (3) the maximum young count in the nest (dead + alive live), (4) the number of fledges recorded for each attempt. From these summaries, the function calculates the maximum number of observed (1) host eggs, (2) young (alive/dead) + unhatched eggs, (3) fledged + unhatched eggs, to get the maximum contents of the nest at each state (lay/incubation, nestling, and fledging)

Depending on the data available, the certainty of the clutch size estimate may vary, or in some cases, may not be estimable. For instance, a successful nest with 4 eggs recorded on several checks and 4 fledged young observed would have high certainty that the clutch size was 4. However, if a nest was observed with 2 eggs on one check and a week later was found predated, the certainty of the clutch size is not known. In this case the clutch size would be estimated as 2 but may have been higher depending on the exact (unobserved) failure date

```{r est clutch}
# Load wrens example data
wrens <- nestwatchR::wren_quickstart
length(unique(wrens$Attempt.ID))
# > [1] 7626    # total attempts in wrens

# How many attempts have un-entered clutch sizes
empty_clutchsize <- wrens %>% filter(is.na(Clutch.Size))
length(unique(empty_clutchsize$Attempt.ID))
# > [1] 2228    # number of NA clutch sizes

# Run estclutchsize
nw.estclutchsize(data = wrens, output = "estimated.clutch")
still_empty <- estimated.clutch %>% filter(is.na(Clutch.Size))
length(unique(still_empty$Attempt.ID))
# > [1] 376     # number of clutch sizes not estimated
```
## Estimate First Lay Date
First lay dates can be estimated if we know how many eggs were present if teh nest was visited during the egg laying phase. The `nw.estfirstlay()` can help a user identify which attempts have enough data to estimate lay dates for. This function takes a merged NestWatch dataframe as `data =`, a user-provided nest phenology dataframe (`phenology = `, see below for details), and has an optional `output =` argument to rename the resulting dataframe. 

This function uses data spread across several datafields within the NestWatch dataset to estimate the lay date fo the first egg of the attempt's clutch. In order, the function determines if (1) the nest was visited during the egg laying phase, (2) the attempt has a hatch date, and (3) if the nest has a fledge date. By checking in this order, the function estimates the lay date from the most proximal event date. If the nest was visited during lay, the estimated lay date will likely be more precise than if estimated off of a fledge date (as there can be variation in incubation and nestling timeframes). By counting backwards in time from the most proximal event through the appropriate nest phases, we can estimate first lay dates. The timeframe between these events (lay, incubation, nestling, fledge), varies by species and may vary due to other variables like latitude. To account for this, this function takes the required `phenology = ` argument, which is a user-provided dataframe with species-level nest phenology data:

```{r phen table, echo = T}
phenology <- data.frame(Species = c("bewwre", "carwre"),
                        Clutch.Size  = c(5, 4),
                        Eggs.per.Day = c(1, 1),
                        Incubation   = c(16, 16),
                        Nestling     = c(16, 13),
                        Total.Nesting.Period = c(50, 40))

DT::datatable(phenology, options = list(paging = F, scrollX = T, searching  = F))
```



