---
title: "Filter NestWatch Data on Finer Scales"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filter NestWatch Data on Finer Scales}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      collapse = TRUE,
                      comment = "#>",
                      out.width = "100%",
                      fig.height = 4, 
                      fig.width = 7, 
                      fig.align = "center")
# only build vignettes locally and not for R CMD check
knitr::opts_chunk$set(eval = nzchar(Sys.getenv("BUILD_VIGNETTES")))
library(nestwatchR)
```

Before analysis, users should consider conducting finer-scale filtering in order to clean the NestWatch dataset after running `nw.cleandata`. This may include selecting certain species, identifying specific nest phenology dates (ie. incubation should not last longer than X days for species Y), or limiting nest attempts to a certain geographic area. 

## Filter Species

Limiting the dataset to just a few species can easily be done using the pipe (`%>%`). If you are unfamiliar with "piping", see the `migritrr` package. Below we will subset the `merged.data` dataframe produced in the Intro vignette to include only attempts for Carolina Wren ("carwre") and Bewick's Wren ("bewwre").

```{r filter to spp, eval = F, echo = T}
wrens <- merged.data %>% filter(Species.Code %in% c("carwre", "bewwre"))

```

